Containers and R---the Rockerverse and beyond

### 2020-01-16 manuscript.Rmd
minor updates to links, fix footnote

**Microsoft R Open** ([MRO](https://github.com/nuest/mro-docker)) is an R distribution formerly known as Revolution R Open (RRO) before Revolution Analytics was acquired by Microsoft.
MRO is compatible with main R and its packages.
"It includes additional capabilities for improved performance, reproducibility, and platform support." \citep{microsoft_mro}
Most notably these capabilities are the MRAN repository, which is enabled by default, and the (optional) integration with [Intel® Math Kernel Library](https://software.intel.com/en-us/mkl) (MKL) for multi-threading in linear algebra operations \citep{microsoft_multithread}.
MRO does not provide official Docker images, but a set of community-maintained `Dockerfile`s and Docker images `nuest/mro` are available on Docker&nbsp;Hub at [https://github.com/nuest/mro-docker](https://github.com/nuest/mro-docker) and on GitHub at [https://github.com/nuest/mro-docker](https://github.com/nuest/mro-docker) respectively.
The images are inspired by the Rocker images and can be used much in the same fashion, effectively a drop-in replacement allowing users to quickly evaluate if the benefits of MRO + Intel® MKL apply to their use case.
Version-tagged images are provided for the latest bugfix release of recent R versions.
Extended license information about MKL is printed at every startup.

@@

**[pqR](http://www.pqr-project.org/)** is a "a pretty quick version of R".
_pqR_ attempts to improve R on some opinionated issues in the R language and is the basis for experimental features, e.g., automatic diffentiation\footnote{\href{https://riotworkshop.github.io/abstracts/riot-2019-pqr.txt}{https://riotworkshop.github.io/abstracts/riot-2019-pqr.txt}}.
The source code development [on GitHub](https://github.com/radfordneal/pqR/) is a one man project and it does not provide any Docker images.
But especially disruptive approaches may contribute to the development of the R ecosystem, so the `nuest/pqr` images on Docker&nbsp;Hub and GitHub at [https://hub.docker.com/r/nuest/pqr/](https://hub.docker.com/r/nuest/pqr/) and [https://github.com/nuest/pqr-docker](https://github.com/nuest/pqr-docker) respectively.

**FastR** is "A high-performance implementation of the R programming language, built on GraalVM" (https://github.com/oracle/fastr).
It is developed by Oracle, connects R to the GraalVM ecosystem \citep{wikipedia_graalvm_2019}, and also claims superior performance but also targets full compatibility with base R\footnote{\href{https://github.com/oracle/fastr}{https://github.com/oracle/fastr}}.
There are no offical Docker images provided, but `nuest/fastr` images and `Dockerfile` are independently maintained on [Docker&nbsp;Hub](https://cloud.docker.com/repository/docker/nuest/fastr/) and [GitHub](https://github.com/nuest/fastr-docker) respectively.

@@

@@

## Using Docker in a commercial Data Science Platform
\label{rstudio}

@@

## Processing [div]
\label{processing}

@@

First, such containers can greatly facilitate **fixing bugs**, because developers can readily start and later discard a container to investigate a bug report without affecting their regular system, and potentially having to revert changes back.
Using the Rocker images with RStudio, these disposable environments lack no development comfort (cf. Section&nbsp;\nameref{compendia}).
\citet{eddelbuettel_debugging_2019} describes an example how a Docker container was used to debug an issue with a package only occuring with a particular version of Fortran, and using tools which are not readily available on all platforms (e.g., not on macOS).
For the [R-spatial community](https://www.r-spatial.org/), using containers is particularly useful, because of the strong integration of system libraries in core packages for geospatial data modelling and analysis.
Purpose built Docker are used to prepare for upcoming releases of system libraries, individual bug reports, and for the lowest supported versions of system libraries \footnote{\href{https://github.com/r-spatial/sf/tree/master/inst/docker}{https://github.com/r-spatial/sf/tree/master/inst/docker}, 
\href{https://github.com/Nowosad/rspatial_proj6}{https://github.com/Nowosad/rspatial_proj6}, \href{https://github.com/r-spatial/sf/issues/1231#issuecomment-570076440}{https://github.com/r-spatial/sf/issues/1231\#issuecomment-570076440}}.

@@

New features, which make complex workflows accessible and reproducible, and variety in packages, even when they have overlapping features, are a signal and a support for a growing user base.
This growth is possibly the most important goal for the foreseeable future in the _Rockerverse_, and just like the Rocker images have matured over years of use and millions of runs the new ideas and prototypes will have to proof themselves.
It should be noted that the dominant position of Docker is a blessing and a curse for these goals.
It could be wise to start experimenting with non-Docker containerisation tools now, e.g. R packages interfacing with other container engines such as [podman/buildah](https://github.com/containers/libpod) or [rkt](https://coreos.com/rkt/), or an R package for creating Singularity image recipe files.
Such efforts help to avoid lock-in and to design sustainable workflows based on concepts of _containerisation_, not on their implementation in Docker.
If adoption of containerisation and R continue to grow, the missing pieces for a success predominantly lie in (a) coordination and documentation of activities to reduce repeated work in favour of open collaboration, (b) the sharing of lessons learned from use cases to build common knowledge and language, and (c) a sustainable continuation and funding for all of development, community support, and education.
A concrete effort to work towards these pieces is to sustain the structure and captured status quo from this work in form of a _CRAN Task View on containerization_.

>>>>

**Microsoft R Open** (MRO) is an R distribution formerly known as Revolution R Open (RRO) before Revolution Analytics was acquired by Microsoft.
MRO is compatible with main R and its packages and it "[..] includes additional capabilities for improved performance, reproducibility, and platform support." \citep{microsoft_mro}
Most notably these capabilities are the MRAN repository, which is enabled by default, and the (optional) integration with [Intel® Math Kernel Library](https://software.intel.com/en-us/mkl) (MKL) for multi-threading in linear algebra operations \citep{microsoft_multithread}.
MRO does not provide official Docker images, but a set of community-maintained `Dockerfile`s and Docker images are provided by the poject `mro-docker` ([https://github.com/nuest/mro-docker](https://github.com/nuest/mro-docker)), e.g. the image [`nuest/mro`](https://hub.docker.com/repository/docker/nuest/mro/).
The images are inspired by the Rocker images and can be used much in the same fashion, effectively a drop-in replacement allowing users to quickly evaluate if the benefits of MRO + Intel® MKL apply to their use case.
Version-tagged images are provided for the latest bugfix release of recent R versions.
Extended license information about MKL is printed at every startup.

@@

**pqR** ([http://www.pqr-project.org/](http://www.pqr-project.org/)) is a "a pretty quick version of R".
_pqR_ attempts to improve R on some opinionated issues in the R language and is the basis for experimental features, e.g., automatic diffentiation\footnote{\href{https://riotworkshop.github.io/abstracts/riot-2019-pqr.txt}{https://riotworkshop.github.io/abstracts/riot-2019-pqr.txt}}.
The source code development [on GitHub](https://github.com/radfordneal/pqR/) is a one man project and it does not provide any Docker images.
But especially disruptive approaches may contribute to the development of the R ecosystem, so the `nuest/pqr` image was independently created, see Docker Hub at [https://hub.docker.com/r/nuest/pqr/](https://hub.docker.com/r/nuest/pqr/) and GitHub at [https://github.com/nuest/pqr-docker](https://github.com/nuest/pqr-docker).

**FastR** ([https://github.com/oracle/fastr](https://github.com/oracle/fastr)) is "A high-performance implementation of the R programming language, built on GraalVM" \citep{oracle_labs_oraclefastr_2020}.
It is developed by Oracle, connects R to the GraalVM ecosystem \citep{wikipedia_graalvm_2019}, and also claims superior performance but also targets full compatibility with base R \citep{oracle_labs_oraclefastr_2020}.
There are no offical Docker images provided, but indepedently maintained experimental images and `Dockerfile`s are provided by [nuest/fastr-docker](https://github.com/nuest/fastr-docker), e.g. [`nuest/fastr` on Docker Hub](https://cloud.docker.com/repository/docker/nuest/fastr/).

@@

_version pinning_ with packages \pkg{remotes} and \pkg{versions} for R, and with system package managers for distros

@@

## Using Docker in a commercial data science platform
\label{rstudio}

@@

## Processing
\label{processing}

@@

First, such containers can greatly facilitate **fixing bugs**, because developers can readily start and later discard a container to investigate a bug report without affecting their regular system, and potentially having to revert changes back.
Using the Rocker images with RStudio, these disposable environments lack no development comfort (cf. Section&nbsp;\nameref{compendia}).
\citet{eddelbuettel_debugging_2019} describes an example how a Docker container was used to debug an issue with a package only occuring with a particular version of Fortran, and using tools which are not readily available on all platforms (e.g., not on macOS).
For the [R-spatial community](https://www.r-spatial.org/), using containers is particularly useful, because of the strong integration of system libraries in core packages for geospatial data modelling and analysis.
Purpose built Docker are used to prepare for upcoming releases of system libraries, individual bug reports, and for the lowest supported versions of system libraries\footnote{Cf. \href{https://github.com/r-spatial/sf/tree/master/inst/docker}{https://github.com/r-spatial/sf/tree/master/inst/docker}, \href{https://github.com/Nowosad/rspatial_proj6}{https://github.com/Nowosad/rspatial\_proj6}, and \href{https://github.com/r-spatial/sf/issues/1231}{https://github.com/r-spatial/sf/issues/1231}}.

@@

New features, which make complex workflows accessible and reproducible, and the variety in packages connected with containerisation, even when they have overlapping features, are a signal and a support for a growing user base.
This growth is possibly the most important goal for the foreseeable future in the _Rockerverse_, and just like the Rocker images have matured over years of use and millions of runs the new ideas and prototypes will have to proof themselves.
It should be noted that the dominant position of Docker is a blessing and a curse for these goals.
It could be wise to start experimenting with non-Docker containerisation tools now, e.g. R packages interfacing with other container engines such as [podman/buildah](https://github.com/containers/libpod) or [rkt](https://coreos.com/rkt/), or an R package for creating Singularity image recipe files.
Such efforts help to avoid lock-in and to design sustainable workflows based on concepts of _containerisation_, not on their implementation in Docker.
If adoption of containerisation and R continue to grow, the missing pieces for a success predominantly lie in (a) coordination and documentation of activities to reduce repeated work in favour of open collaboration, (b) the sharing of lessons learned from use cases to build common knowledge and language, and (c) a sustainable continuation and funding for all of development, community support, and education.
A concrete effort to work towards these pieces is to sustain the structure and captured status quo from this work in form of a _CRAN Task View on containerization_.
